---
title: "Mixed Model Estimation via Maximum Likelihood, <br> and a Connection to Additive Models"
output: 
  html_document: 
    keep_md: yes
    theme: united
    toc: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message=F)
```



## Introduction
This is an example based on Wood, 2006, chapter 6 in particular.  It assumes familiarity with standard regression from a matrix perspective. 

## Model Formulation

We can start with a standard linear model expressed as follows:

$$\mathbf{y} = \mathbf{Xb} + \mathbf{\epsilon} $$

Here $\mathbf{y}$ is the target variable, $\mathbf{X}$ is a model matrix (first column representing the intercept, the rest are the covariates of interest), $\mathbf{b}$ are the coefficients, and error $\mathbf{\epsilon}$. Note that beyond this point I'll largely refrain from using bold to indicate vectors/matrices, or using subscripts for every *i*<sup>th</sup> observation. Let's just assume we are in a typical data situation involving multiple observations, a univariate vector target variable (y), a matrix of predictor variables (X) etc.

For a mixed model with a single random effect for some grouping factor (e.g. students within schools), this extends to:

$$y = Xb + Zg + \epsilon$$

Where Z is an indicator matrix pertaining to the grouping structure (sometimes referred to as dummy coding or one-hot encoding). Consider a factor z representing group/cluster membership, this would convert z to the following:

```{r dummycode, echo=FALSE}
z = Z = factor(c('A','A','B','B','C','C'))
Z =  model.matrix(~ Z-1)
pander::pander(data.frame(z, Z))
```

The coefficients $g$ are the random effects, assumed $\mathcal{N}(0,\tau)$, and while we are often interested in them, they do not have to be estimated directly.

$$y = Xb + Zg + \epsilon \\
g \sim \mathcal{N}(0, \psi_\theta) \\
\epsilon \sim \mathcal{N}(0, \Lambda\sigma^2)$$

In this depiction $\psi_\theta$ can reflect some more interesting dependencies, but in the simple case of a random intercepts model it can be a single variance estimate $\tau^2$. $\Lambda$ can be used to model residual covariance but often is just the identity matrix, with the underlying assumption of constant variance $\sigma^2$ across observations.  

We can combine the random and residuals into a single construct reflecting the covariance structure of the observations:

$$ e = Zg + \epsilon $$

This makes $\mathbf{e}$ a multivariate vector with mean 0 and covariance (**I** is the unit matrix):

$$Z\psi_{\theta}Z^\intercal + I\sigma^2$$

This puts us back to a standard linear model:

$$ y = Xb + e, \\
e \sim \mathcal{N}(0, \Sigma_\theta\sigma^2)$$


## Maximum Likelihood Estimation

Given where we are now, we can proceed to estimate the mixed model. For this we'll use the sleepstudy data from lme4. The data has reaction times for 18 individuals over 10 days each (see the help file for the sleepstudy object for more details).

### Data
```{r dataSetup}
data(sleepstudy, package='lme4')
X = model.matrix(~Days, sleepstudy)
Z = model.matrix(~factor(sleepstudy$Subject)-1)
y = sleepstudy$Reaction
```

### ML function
The following is based on the code in Wood (6.2.2), with a couple modifications for consistent nomenclature. $\theta$ represents the vector of parameters we wish to estimate. The (square root of the) variances will be estimated on the log scale. In Wood, he simply extracts the 'fixed effects' for the intercept and days effects using lm (6.2.3).

```{r mlfunc}
llMixed = function(y, X, Z, theta){
  tau = exp(theta[1])
  sigma = exp(theta[2])
  n = length(y)
  
  # evaluate cov mat for y
  e = tcrossprod(Z)*tau^2 + diag(n)*sigma^2
  L = chol(e)  # L'L = e
  
  # transform dependent linear model to independent
  y = backsolve(L, y, transpose=TRUE)
  X = backsolve(L, X, transpose=TRUE)
  b = coef(lm(y~X-1))
  LP = X %*% b
  
  ll = -n/2*log(2*pi) -sum(log(diag(L))) - crossprod(y-LP)/2
  -ll
}
```


Here is an alternative function using a multivariate approach that doesn't use the transformation to independent, and might provide additional perspective..

```{r mlfuncMV}
llMixedMV = function(y, X, Z, theta){
  tau = exp(theta[1])
  sigma = exp(theta[2])
  n = length(y)
  
  # evaluate cov mat for y
  e = tcrossprod(Z)*tau^2 + diag(n)*sigma^2

  b = coef(lm.fit(X, y))
  mu = X %*% b

  ll = -mvtnorm::dmvnorm(y, mu, e, log=T)
}
```



### Results

We'll use the optim function for estimation.  A slight change to tolerance is included to get a closer estimate to lme4.

```{r optim}
paramInit = c(0, 0)
names(paramInit) = c('tau', 'sigma')

modelResults = optim(llMixed, X=X, y=y, Z=Z, par=paramInit, control=list(reltol=1e-10))
modelResultsMV = optim(llMixedMV, X=X, y=y, Z=Z, par=paramInit, control=list(reltol=1e-10))

rbind(c(exp(modelResults$par), logLik = modelResults$value, coef(lm(y~X-1))),
      c(exp(modelResultsMV$par), logLik = modelResultsMV$value, coef(lm(y~X-1)))) %>% 
  round(2)
```

As we can see, both formulations produce identical results. We can now compare those results to the lme4 output for the same model.

```{r lme}
library(lme4)
lmeMod = lmer(Reaction ~ Days + (1|Subject), sleepstudy, REML=FALSE)
lmeMod
```

We can predict the random effects (Wood, 6.2.4), and after doing so compare to the lme4 estimates.

```{r estRanEf}
tau = exp(modelResults$par)[1]
tausq = tau^2
sigma = exp(modelResults$par)[2]
sigmasq = sigma^2
Sigma = tcrossprod(Z)*tausq/sigmasq + diag(length(y))
ranefEstimated = tausq*t(Z)%*%solve(Sigma) %*% resid(lm(y~X-1))/sigmasq
data.frame(ranefEstimated, lme4 = ranef(lmeMod)$Subject[[1]]) %>% round(2)
```


## Additive model as a mixed model 

At this point I'd like to demonstrate some concepts from section 6.6 in Wood. Conceptually, the take home idea is that an additive model, or generalized additive model (GAM), can be seen as a mixed model, which is interesting in and of itself (at least to me), but it also means that GAMs meld nicely with mixed models generally.  For an intro on additive models, one can see my [document](https://sites.google.com/a/umich.edu/micl/miscfiles/GAMS.pdf), which is more or less an overview of Wood's text.

### Data set up

See Wood 3.2.  The data regards motor engine size and wear in 19 Volvos, with the initial assumption that larger capacity engines will wear out less quickly.

```{r gamData}
size = c(1.42,1.58,1.78,1.99,1.99,1.99,2.13,2.13,2.13,2.32,2.32,2.32,2.32,2.32,2.43,2.43,2.78,2.98,2.98)
wear = c(4.0,4.2,2.5,2.6,2.8,2.4,3.2,2.4,2.6,4.8,2.9,3.8,3.0,2.7,3.1,3.3,3.0,2.8,1.7)

x = size - min(size)
x = x / max(x)
d = data.frame(wear, x)
```


### Relevant functions
We'll create functions for the cubic spline operation, the creation of a model matrix, the creation of the penalty matrix, and finally the fitting function.

```{r gamFuncs}
# cubic spline function
rk <- function(x, z) {
  ((z-0.5)^2 - 1/12) * ((x-0.5)^2 - 1/12) / 4 -
    ((abs(x-z)-0.5)^4 - (abs(x-z)-0.5)^2/2 + 7/240) / 24
}

# create the model matrix
splineX <- function(x, knots) {
  q <- length(knots) + 2 # number of parameters
  n <- length(x) # number of observations
  X <- matrix(1, n, q) # initialized model matrix
  X[, 2] <- x # set second column to x
  X[, 3:q] <- outer(x, knots, FUN = rk) # remaining to cubic spline
  X
}

# set up the penalized regression spline penalty matrix, given knot sequence xk
Sfunc = function(xk){
  q = length(xk)+2
  S = matrix(0, q, q) # initialize
  S[3:q, 3:q] = outer(xk, xk, FUN=rk)
  S
}

# Matrix sqrt function
matSqrt = function(S){
  UDU = eigen(S, symmetric=TRUE)
  U = UDU$vectors
  D = diag(UDU$values)
  B = crossprod(U) %*% sqrt(D)
  B
}

# the fitting function
prsFit <- function(y, x, knots, lambda) {
  q = length(knots) + 2 # dimension of basis
  n = length(x) # number of observations
  Xa = rbind(splineX(x, knots), matSqrt(Sfunc(knots)) * sqrt(lambda)) # augmented model matrix
  y[(n + 1):(n + q)] = 0 #augment the data vector
  lm(y ~ Xa - 1) # fit and return penalized regression spline
}
```

### Fit a penalized model

Now we can fit the model and visualize.

```{r fitCSgam}
xk = 1:7/8  # choose some knots
mod = prsFit(y=wear, x=x, knots=xk, lambda=.0001) # fit the penalized spline

xp = 0:100/100  # values for prediction
Xp = splineX(xp, xk)

plot(x, wear, xlab='Scaled Engine size', ylab='Wear Index', pch=19,
     col="#FF5503", cex=.75, col.axis='gray50', bty='n')
lines(xp,Xp%*%coef(mod), col='#2957FF') 
```

### As a mixed model

One can use the result of eigen decomposition on the penalty matrix to ultimately produce a re-parameterization of the original matrix as fixed and random effects components.

```{r gam2mixed}
S = Sfunc(xk)
init = eigen(S)
U = init$vectors
D = diag(init$values)
poseigen = which(diag(D) > 0)  
Dpos = D[poseigen, poseigen]           # smallest submatrix containing all positive values
Xf = splineX(x, knots = xk)            # spline model matrix
U_F = U[, (ncol(U)-1):ncol(U)]         # partition eigenvector matrix
U_R = U[, 1:(ncol(U)-ncol(U_F))]
X_F = Xf %*% U_F                       # fixed part  with B_F coef to be estimated (not penalized)
X_R = Xf %*% U_R                       # random part with B_R random effects
Z = X_R %*% sqrt(Dpos)
```

The above operations have effectively split the GAM into fixed and random parts:

$$\mathbf{X}_F\mathbf{\beta}_F + \mathbf{X}_R\mathbf{b}_R, \\
\mathbf{b}_R\sim \mathcal{N}(\mathbf{0}, \mathbf{D_+^{-1}}/\lambda) \\$$

Here $\lambda$ is a smoothing parameter, which controls the amount of smoothing. For a penalized spline, the loss function is:

$$\lVert y - X\beta \rVert^2 + \lambda\beta^{\intercal}S\beta ,$$

with the second part the added penalty. As $\lambda$ approaches infinity, we essentially get straight line fit, while a $\lambda$ of 0 would be the same as an unpenalized fit.


The Z above represents part of the mixed model Z we had before in the standard mixed model. We can represent it as follows:

$$\mathbf{X}_F\mathbf{\beta}_F + \mathbf{Zg}, \\
\mathbf{g} \sim \mathcal{N}(\mathbf{0}, \mathbf{I}/\lambda) \\$$


To incorporate a gamm, i.e. a *generalized additive mixed model*, the X_F above would become part of the 'fixed' effect part of the model, while Z would be part of the random effects, and estimation would proceed normally as for a mixed model.  


Initially we'll just duplicate a standard mixed model using the mgcv package.

```{r gamSleepStudy}
library(mgcv); library(gamm4)
sleepstudy$Subject = factor(sleepstudy$Subject)
modGam = gamm4(Reaction ~ Days, random=~(1|Subject), data=sleepstudy)
summary(modGam$mer)
```


Now we'll add a cubic spline for the effect of Days.  We can see the smooth term listed as random effect.
```{r}
modGamS = gamm4(Reaction ~ s(Days, bs='cs'), random=~(1|Subject), data=sleepstudy)
summary(modGamS$mer)    
# summary(modGamS$gam)
plot(modGamS$gam)

# lmer(Reaction ~ (1+Days|Subject), data=sleepstudy)
```


Another way to look at this is that we have added the capacity to examine nonlinear relationships and other covariance structures to the standard mixed model framework, making a very flexible modeling approach even more so.
